Learning Network Growth Mechanisms from Degree Distributions in Preferential Attachment Networks

### 2025-08-17 paper.tex
Making changes

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-basic,
]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

@@

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Tail Flexibility in the Degrees of Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, discrete extremes, power law, preferential
attachment},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Tail Flexibility in the Degrees of Preferential Attachment
Networks]{Tail Flexibility in the Degrees of Preferential Attachment
Networks}

@@

Networks appear in many fields such as sociology, politics,
epidemiology, and economics. Statistical methods have been used to study
networks that appear in these fields and beyond, from the use of
stochastic block models to detect communities within the French
political blogosphere \citep{Latouche11}, to using Exponential Random
Graph Models (ERGMs) to analyse the global trade network
\citep{Setayesh22}, and the use of mechanistic models in neuroscience to
investigate the patterns of connections in neural systems
\citep{Betzel17}.

One example of mechanistic models is the preferential attachment (PA)
model, where as a network grows and new vertices join, they connect to
those already present in the network with probabilities proportional to
some preference function of their in-degrees. The well-known
Barabási-Albert (BA) model is a special case of this when \(b(k) = k+1\)
\citep{Barabasi99}.

These mechanistic models are generally useful when simulating networks,
but not for fitting to real data as usually only snapshots of a network
are available and not the full evolution - without this conclusions
about the growth are difficult to make. However, one statistic that is
usually available for snapshots of networks is the degree distribution,
making it valuable to study the degrees in a network. Often, it is
appealing to model the degree distributions of real networks using the
power law \footnote{Power law: \(\bar F(k) = L(k)\times k^{-1/\xi}\),
  where \(\xi>0\) and \(L(k)\) is slowly varying in \(k\).}, as it is
seemingly observed in much of the data. Furthermore, as the BA model
results in a power law degree distribution \citep{Barabasi99}, that the
power law fits well offers the interpretation that the network may have
grown according to PA or similar simple rules. However, attempting to
explain the networks growth using preferential attachment when the
degrees seemingly follow a power law may not be the best course of
action as \citet{krapivsky01} show that a general preference function
does not necessarily lead to power law degrees. Specifically, showing
that a preference function of form \(b(k) = k^\alpha\) leads to light
tails when \(\alpha<1\), a degenerate network (a finite number of
vertices end up with all of the added edges) when \(\alpha>1\), and only
yields heavy tails when \(\alpha=1\). \citet{rudas07} derives the
limiting degree distribution in terms of the preference function
\(b(\cdot)\) and its parameters, we build upon this by finding a class
of preference functions that result in realistic tails i.e.~heavy but
not as heavy as implied by the power law. On the other hand, it is still
being argued whether or not many real data actually follow the power
law. \citet{Broido_2019} provides some evidence that real world networks
rarely have degrees that follow a power law, through the use of
statistical analysis of almost 1000 networks across a wide variety of
fields; \citet{Voitalov_2019}, on the other hand, disagrees and claims
that these kinds of networks are not nearly as rare and only appear so
as a result of an unrealistic expectation of power law without
deviations or noise.

Extreme value methods have been scarcely used when it comes to analysis
of degree distributions and whether they follow the power law. The
absence of these methods essentially downweights the deviation of
vertices with the largest degrees, which are usually the most
influential, from the power law. \citet{Voitalov_2019} proposes
estimating the power law exponent of a degree distribution using three
estimators based in extreme value theory, in contrast to the estimators
generally used in network science, the degree distributions are then
split into one of three categories (not power-law, hardly power-law or
power-law) depending on the values of the estimates obtained from each
of the estimators. \citet{Lee24} proposes a framework for modelling
entire degree distributions using a combination of two generalisations
of the power law with a model selection step that is capable of
suggesting whether or not the data is adequately modelled by a power
law, with one conclusion being that most data seems to have a tail
lighter than what is implied by the power law in body whilst still being
heavy tailed. One of the models used by \citet{Lee24} is the Zipf-IGP
model where up until a certain threshold the data is modelled using the
Zipf disribution (a discrete power law) and beyond it a discrete version
of the Generalised Pareto distribution is used, we will later compare
our model to this as they share some similarities.
\citet{wang2022random} considers a preferential attachment model with an
affine preference function and reciprocal edges, finding that the joint
limiting distribution of the in-degrees and out-degrees is multivariate
regularly varying and has the property of hidden regular variation.

Evidently, the study of degree distributions is a very active but the
vast majority of research in the field (whether using extremes or not)
are purely descriptive in the sense that no information about the
preference function is revealed even if preferential attachment was the
underlying mechanism. This paper aims to address this gap, asking if the
degree distribution of a network is assumed to come from the
preferential attachment model, can we directly infer the model
parameters?

The remainder of this paper is as follows: Section 2 gives a detailed
description of the PA model alongside the theoretical results for the
limiting degree distribution with a focus on its tail behaviour in terms
of the preference function \(b(\cdot)\). It is shown that a heavy tailed
degree distribution can only be the result of an asymptotically linear
preference function ending with a suggestion for a preference function
that can guarantee heavy tails while remaining flexible.
Section~\ref{sec-model} utilises the preference function proposed at the
end of Section 2, and illustrates how the tail index of the degree
distribution varies with the model parameters. Section~\ref{sec-sim}
consists of a simulation study where networks are simulated from the
proposed model for various parameter combinations, demonstrating that
the parameters can be recovered from fitting the model to only the
degree distribution. Section~\ref{sec-real} fits the model to some real
data and provides posterior estimates for the preference function.
Section~\ref{sec-conc} concludes the article.

@@

The network generative model that we will focus on in this paper is the
General Preferential Attachment model in \citet{rudas07} and is defined
as follows:

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+\varepsilon\), which leads to a power-law degree distribution
with \(\xi=1/2\), and the Uniform Attachment (UA) model where \(b(k)=c\)
leading to a degree distribution that is light tailed.

Given condition \ref{eq-condb2} the survival function of the limiting
degree distribution, called the limiting survival hereafter, can be
analytically derived in the case where \(m=1\), obtained by considering
a branching process that is equivalent to the growth of the network, as
in \citet{rudas07}. Theorem 1 from \citet{rudas07} states that for the
tree \(\Upsilon(t)\) at time \(t\):

@@

We are interested in how the tail behaviour of the discrete limiting
degree distribution is affected by the preference function \(b\), which
can be determined by the results of \citet{shimura12}.

@@

\citet{shimura12} states that if
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy-tailed with \(\bar F(k) \sim k^{-\alpha}\). On the
other hand, if \(\lim_{k\rightarrow\infty} \Omega(F,k) = 0\) then the
distribution is light-tailed. This allows us to show the following:

@@

Proposition~\ref{prp-omega} aligns with the result from
\citet{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light-tailed distribution, as
\(\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0\) if \(b(k)=k^\alpha\) where
\(\alpha < 1\). Proposition 2.1 also aligns with the fact that BA model
produces a heavy-tailed degree distribution with index \(\xi=0.5\) by
considering the preference function \(b(k) = k + \varepsilon\), as
\(\lim_{k\rightarrow\infty}b(k+1)-b(k)=1\) leaving the tail index to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\). So,
in order for the degree distribution to be heavy-tailed we need that the
limit \(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is positive.
To determine the class of functions that will result in heavy-tailed
limiting degree distributions, we use the following result:

@@

Consider a GPA model with preference function \(b(\cdot)\) satisfying
Proposition~\ref{prp-omega}. Then the limiting degree distribution has a
heavy tail with index \(c/\lambda^{*}\) if and only if
\(\lim_{k\rightarrow\infty}b(k)/k = c > 0\).

@@

\[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
\] meaning by Theorem~\ref{thm-stolz}:

@@

Considering Proposition~\ref{prp-omega2} we propose the following
preference function:

@@

This preference function, as per Proposition~\ref{prp-omega}, will
produce a degree distribution with tail index \(\beta/\lambda^*\)
guaranteeing a heavy tail. We study this preference function further in
the next section.

@@

\section{Preferential Attachment with Flexible Heavy
Tail}\label{sec-model}

In the previous section, we found that using a preference function with
linearity in the limit allows for the inclusion of sub/super-linear
behaviour below the threshold, while simultaneously guaranteeing a heavy
tail. In this section, we study how the limiting degree distribution
varies with the preference function according to Equation~\ref{eq-pref}.
Its survival is

@@

\begin{verbatim}
     a   b  eps k0 x   lambda        surv
10 0.5 1.5 0.01 20 0 1.500094 0.006622105
\end{verbatim}

\begin{verbatim}
[1] 0.9999374
\end{verbatim}

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\) ensures that
the shape parameter of the IGP is positive and thus the distribution is
heavy-tailed. Additionally the value of the shape parameter \(\xi\) is
shown in Figure~\ref{fig-polyheat} for various parameter choices. The
darker and lighter regions on the heat maps correspond to a heavier and
a lighter tail, respectively, the red dashed line shows combinations of
\(\alpha\) and \(\beta\) that produce a limiting degree distribution
with the same tail index as the Barabási-Albert model, that is,
\(\xi=0.5\).

@@

This first subsection aims to show that the parameters of the model in
Section~\ref{sec-model} can be recovered from simulating a network from
the model, and fitting it to the observed degree distribution, using the
likelihood in \eqref{eq-lh}.

@@

\begin{align*}
\alpha&\sim \text{Ga}(1,0.01),\\
\beta &\sim  \text{Ga}(1,0.01),\\
k_0 &\sim \text{U}(1,10,000),\\
\varepsilon &\sim \text{Ga(1,0.01)},
\end{align*}

where Ga(a, b) is the Gamma distribution with shape a and rate b, and
U(a, b) is uniform distribution with lower and upper bounds a and b, to
obtain a posterior distribution, up to the proportionality constant.
Posterior samples can then be obtained by an adaptive
Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm. For these
simulated networks \(l=0\).

@@

Figure~\ref{fig-rec1} and Figure~\ref{fig-rec2} shows the estimations
demonstrates the usefulness of the model, as we can recover the model
parameters fairly well from only the final degree distribution of the
simulated network. This indicates that the method may also be applied to
the degree distributions of real networks, estimating the model
parameters assuming they evolved according to the GPA scheme.

@@

Figure~\ref{fig-real1} displays the posterior estimates of the survival
function for various data sets, obtained from fitting the GPA model and
the Zipf-IGP mixture model from \citet{Lee24}. In most cases, the GPA
model does not necessarily provide an improvement in fit when compared
to the Zipf-IGP model but where the GPA model fits well we gain
additional information about the preference function assuming that the
network evolved according the the GPA scheme.

@@

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-basic,
]{sn-jnl}

\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Learning Network Growth Mechanisms from Degree Distributions in Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, discrete extremes, power law, preferential
attachment},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title[Learning Network Growth Mechanisms from Degree Distributions in
Preferential Attachment Networks]{Learning Network Growth Mechanisms
from Degree Distributions in Preferential Attachment Networks}

@@

Networks have become very powerful tools for representing and analysing
complex systems, with uses in a large array of fields, from using
stochastic block models to detect communities online \citep{Latouche11},
to using Exponential Random Graph Models (ERGMs) to analyse the global
trade network \citep{Setayesh22}, and even using mechanistic models to
investigate the patterns in neural systems \citep{Betzel17}.

Along with the recent rise of interest in networks, has been a debate
amongst those that study them regarding the claims that most real
networks are scale-free. This claim is most often equivalent to claiming
that the degree distributions of real networks follow a power law, that
is the fraction of nodes with degree \(k\) is proportional to
\(k^{-\alpha}\), and therefore has a regularly varying tail with extreme
value index \(1/\alpha\). \citet{Broido_2019} provide empirical
evidence, using almost one thousand networks, that these scale-free
networks do not make up a large majority of real networks. They compare
the fits of a power law model against that of several non-scale-free
models only finding strong evidence for scale-freeness in four percent
and weak evidence in fifty-two percent of the networks. On the other
side of this debate is \citet{Voitalov_2019} who disagrees and claims
that these networks are not nearly as rare and only appear to so as a
result of an unrealistic expectation of a power law without deviations
or noise. Evidence of these deviations from a power law is shown by
\citet{Lee24} who demonstrate that a lot of networks are partially
scale-free, in that the body of the degree distribution is modelled well
by a power law while the tail is lighter than what is implied by the
body (although still regularly varying).

Most of the study into the degrees of real networks, including those
mentioned here, has been purely descriptive in the sense that no
information about the networks growth is revealed (aside from relating a
power law to preferential attachment).

The study of network growth began to gain popularity when
\citet{Barabasi99} showed that a network that has grown in such a way
that nodes gain edges at a rate proportional to their degree results in
a network that is scale-free. This is why the presence of real
scale-free networks is such a hot topic, since if it is shown that they
are often scale-free then a model such as preferential attachment gains
a lot of justification for being used to model network growth. The
preferential attachment model they introduced uses a preference
function, \(b(k)=k+1\) where \(k\) is a node in-degree, that is fairly
simple and provided the foundations for many modifications and
extensions to the model. \citet{krapivsky01} considered a more general
preference function of the form \(b(k) = (k+1)^\alpha\) and showed that
the degree distribution only has regular variation in the tail when
\(\alpha=1\), no regular variation when \(0<\alpha<1\) (and therefore
not power law), and when \(\alpha>1\) a finite number of nodes end up
with all edges after a certain point resulting in a degenerate degree
distribution. \citet{wang2022random} returns to a linear preference
function of the form \(b(k) = k+\varepsilon\) but adds the possibility
for reciprocal edges to be sent, this results in joint in-degree
out-degree distribution that is multivariate regular varying and has the
property of hidden regular variation. \citet{rudas07} follows in the
footsteps of \citet{krapivsky01} and considers a preferential attachment
tree with a preference function \(b(\cdot)\) and uses theory from
continuous time branching processes to derive a limiting degree
distribution in terms of the preference function \(b(\cdot)\). Research
in this area tends to only focus on the theoretical asymptotic results
of network growth models with little analysis of real networks. This
paper aims to address this gap, asking if a network is assumed to come
from a preferential attachment model can we use the degree distribution
alone to directly infer the model parameters?

It is important to think about how we intend to consider the tail of the
degree distributions, because if not done correctly we may end up
essentially discounting the effects of the largest degrees (often the
most influential nodes) deviating from a power law. For this reason we
will use methods from discrete extreme value theory, as they best
capture the nature of the data involved, more specifically we will use
results from \citet{shimura12} to study how the tail of the degree
distribution is affected by the preference function since we wish to
align what we do with the recent findings that the a lot of real degree
distributions are regularly varying.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the preferential attachment model alongside the
theoretical results for the survival of the limiting degrees with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). It is shown that a regularly varying degree distribution
can only be the result of an asymptotically linear preference function.
This section also introduces a preference function that can guarantee
regular variation in the degrees while remaining flexible via a set of
parameters that can provide behaviour similar to that in
\citet{krapivsky01} up until a threshold. Section~\ref{sec-model}
utilises the preference function proposed at the end of
Section~\ref{sec-tail}, and illustrates how the extreme value index
(EVI) of the degree distribution varies with the model parameters.
Section~\ref{sec-sim} consists of a simulation study where networks are
simulated from the proposed model for various parameter combinations,
demonstrating that the parameters can be recovered from fitting the
model to only the degree distribution. Section~\ref{sec-real} fits the
model to some real data and provides posterior estimates for the
preference function. Section~\ref{sec-conc} concludes the article.

@@

The model that we will focus on in this paper is the General
Preferential Attachment model in \citet{rudas07} and is defined as
follows:

@@

Special cases of this model include the Barabási-Albert (BA) model when
\(b(k) = k+\varepsilon\), which leads to a power-law degree distribution
with EVI \(\xi=1/2\), and the Uniform Attachment (UA) model where
\(b(k)=c\) leading to a degree distribution that is not regularly
varying.

The survival function of the limiting degree distribution, called the
limiting survival hereafter, under condition \ref{eq-condb2} can be
analytically derived in the case where \(m=1\).

Consider a continuous time branching process \(\zeta(t)\) driven by a
Markovian pure birth process, with \(\zeta(0)=0\) abd birth rates
depending on a non-negative function \(b(\cdot)\):

\[
\Pr(\zeta(t+dt)=k+1|\zeta(t)=k) = b(k)dt + o(dt).
\] Now let \(\Upsilon(t)\) be the tree determined by \(\zeta(t)\) as
follows: \(\Upsilon(t)=\{\emptyset\}\) and \(\Upsilon(t)=G\) where each
existing node \(x\) in \(\Upsilon(t)\) gives birth to a child with rate
\(b(\mathrm{deg}(x, \Upsilon(t))\) independently of the other nodes
where \(\mathrm{deg}(x, \Upsilon(t))\) is the degree of node \(x\) in
the tree \(\Upsilon(t)\) at time \(t\).

Theorem 1 from \citet{rudas07} states that for the tree \(\Upsilon(t)\)
at time \(t\):

@@

We are interested in how the tail behaviour of the discrete limiting
degree distribution is affected by the preference function \(b\).

@@

\citet{shimura12} states that if
\(\lim_{k\rightarrow\infty} \Omega(F,k) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is regularly varying with \(\bar F(k) \sim k^{-\alpha}\). On
the other hand, if \(\lim_{k\rightarrow\infty} \Omega(F,k) = 0\) then we
will refer to the distribution as light-tailed. This allows us to show
the following:

@@

Proposition~\ref{prp-omega} aligns with the result from
\citet{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light-tailed distribution, as
\(\lim_{k\rightarrow\infty} b(k+1)-b(k) = 0\) if \(b(k)=k^\alpha\) where
\(\alpha < 1\). Proposition 2.1 also aligns with the fact that BA model
produces a regularly varying degree distribution with EVI \(\xi=0.5\) by
considering the preference function \(b(k) = k + \varepsilon\), as
\(\lim_{k\rightarrow\infty}b(k+1)-b(k)=1\) leaving the tail index to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\). So,
in order for the degree distribution to be regularly varying we need
that the limit \(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is
positive. To determine the class of functions that will result in
regularly varying limiting degree distributions, we use the following
result:

@@

Consider a GPA model with preference function \(b(\cdot)\) satisfying
Proposition~\ref{prp-omega}. Then the limiting degree distribution is
regularly varying with EVI \(c/\lambda^{*}\) if and only if
\(\lim_{k\rightarrow\infty}b(k)/k = c > 0\).

@@

\[
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
\] meaning by Theorem~\ref{thm-stolz} in Appendix \ref{sec-sup}:

@@

Using this result we can understand how the preference function is
directly connected with the EVI and regular variation of the degree
distribution. We use this result to create a preference function that
guarantees regular variation in the tail of the degree distribution,
aligning with analysis of real networks, whilst allowing for the tail to
deviate from the shape implied by the body. This gives the model the
capability to produce realistic behaviour in the degrees like what was
in \citet{Lee24} by using a piecewise function inspired by the observed
deviation from the power law after a certain threshold.

@@

This preference function, as per Proposition~\ref{prp-omega}, will
produce a degree distribution with EVI \(\xi=\beta/\lambda^*\)
guaranteeing regular variation. We study this preference function
further in the next section.

@@

\section{A Preferential Attachment Model with Flexible Regular
Variation}\label{sec-model}

In the previous section, we found that using a preference function with
linearity in the limit allows for the inclusion of sub/super-linear
behaviour below the threshold, while simultaneously guaranteeing regular
varation of the degrees. In this section, we study how the limiting
degree distribution varies with the preference function according to
Equation~\ref{eq-pref}. Its survival is

@@

@@

In agreement with Proposition~\ref{prp-omega2}, \(\beta>0\) ensures that
the shape parameter of the IGP is positive and thus the distribution is
regularly varying. Additionally the value of the shape parameter \(\xi\)
is shown in Figure~\ref{fig-polyheat} for various parameter choices. The
darker and lighter regions on the heat maps correspond to a heavier and
a lighter tail, respectively, the red dashed line shows combinations of
\(\alpha\) and \(\beta\) that produce a limiting degree distribution
with the same tail index as the Barabási-Albert model, that is,
\(\xi=0.5\).

Through the connection to the IGP, fitting this model is almost
equivalent to fitting the IGP to the degree and estimating the
parameters but instead of only describing the shape of the degree
distribution we would also gain estimates for the shape of the
preference function while can help understand the mechanisms underlying
the growth of the network.

@@

This first subsection aims to show that the parameters of the model (and
therefore the preference function) in Section~\ref{sec-model} can be
recovered from simulating a network from the model, and fitting it to
the observed degree distribution, using the likelihood in \eqref{eq-lh}.

@@

\begin{align*}
\alpha&\sim \text{Gamma}(1,0.01),\\
\beta &\sim  \text{Gamma}(1,0.01),\\
k_0 &\sim \text{U}(1,10,000),\\
\varepsilon &\sim \text{Gamma(1,0.01)},
\end{align*}

where Gamma(a, b) is the Gamma distribution with shape a and rate b, and
U(a, b) is uniform distribution with lower and upper bounds a and b, to
obtain a posterior distribution, up to the proportionality constant.
Posterior samples can then be obtained by an adaptive
Metropolis-Hastings Markov chain Monte Carlo (MCMC) algorithm. For these
simulated networks \(l=0\).

@@

Figure~\ref{fig-rec1} and Figure~\ref{fig-rec2} shows the estimations
demonstrates the usefulness of the model, as we can recover the model
parameters well from only the final degree distribution of the simulated
network. This indicates that the method may also be applied to the
degree distributions of real networks, estimating the model parameters
assuming they evolved according to the GPA scheme.

@@

Figure~\ref{fig-real1} displays the posterior estimates of the survival
function for various data sets, obtained from fitting the GPA model and
the Zipf-IGP mixture model from \citet{Lee24}. In most cases, the GPA
model gives a similar fit to the Zipf-IGP model but where the GPA model
fits well we gain additional information about the preference function
assuming that the network evolved according the the GPA scheme.

@@

\section{Supplementary Results}\label{sec-sup}
